Co-moving stars in \textsl{Gaia DR1}

### 2016-10-14 paper/ms.tex
methods section

% to-do list
% ----------

% style notes
% -----------
% - This file generates by Makefile; don't be typing ``pdflatex'' or some bullshit.
% - Line break between sentences to make the git diffs readable.
% - Use \, as a multiply operator.
% - Reserve () for function arguments; use [] or {} for outer shit.
% - Use \sectionname not Section, \figname not Figure, \documentname not Article or Paper or paper.

@@

% packages
\definecolor{cbblue}{HTML}{3182bd}
\usepackage{microtype}  % ALWAYS!
\usepackage{amsmath}
\hypersetup{backref,breaklinks,colorlinks,urlcolor=cbblue,linkcolor=cbblue,citecolor=black}

@@

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\msun}{{\rm M}_\odot}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}

@@

The abundance of pairs of stars with small velocity difference in the
high-precsision sample of \figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at separations as large
as a few parsecs. Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances.
  \item We assume that the tangential velocities of a given pair of stars in the
    \tgas\ sample (relative to the solar system barycenter) are either (1)
    co-moving with velocities equal to the circular velocity of a $2~\msun$
    binary star system with the given physical separation, (2) co-moving with
    velocities drawn from a zero-mean Gaussian with velocity dispersion $\sigma
    = 1~\kms$, or (3) drawn from a mixture of three isotropic, zero-mean
    Gaussian distributions with velocity dispersions equal to $(\sigma_1,
    \sigma_2, \sigma_3) = (8, 32, 128)~\kms$ meant to represent young disk
    stars, old disk stars, and halo stars.
\end{itemize}

Under these assumptions, the likelihood function for the proper-motion data for
a star given true distance, $d$, and true tangential velocity, $\bs{v}^\perp =
(\begin{array}[t]{c c} v_\alpha & v_\delta\end{array})^\mathsf{T}$ is
\begin{align}
  L(\bs{v}^\perp, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{\mu} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{\mu} - \bs{x}_\theta \right) \right] \\
  \bs{x}_\theta &= d^{-1} \, \bs{v}^\perp
\end{align}

\todo{APW is working on this text...}
We compute the fully
marginalized likelihood (FML) that the pair has the same 3-space velocity
(hypothesis 1, $\mathcal{L}_1$) and the FML of the stars having different
3-space velocities (hypothesis 2, $\mathcal{L}_2$). We use the ratio of these
FMLs as a scalar for selecting candidate co-moving stars, as described below in
more detail. The likelihood functions $L_1, L_2$ are marginalized over true
3-space velocity and distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \label{eq:hyp2}
\end{align}
where

and we compute the posterior pdfs for the true distances, $p(d\given\varpi)$,
using a uniform space density prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{equation}
  p(d) \propto d^2
\end{equation}
We assume Gaussian priors on the true velocity components
\begin{equation}
  p(\bs{v}) = \normal(0,25)~\kms
\end{equation}

For hypothesis 1, the integral over velocity is non-trivial because the same
``true'' 3-space velocity vector projects into tangential velocity components
differently depending on sky position ($v_\alpha = v_\alpha(\bs{v}, \alpha,
\delta)$). The marginalization integral for hypothesis 2 is simpler because the
integral in \eqname~\ref{eq:hyp2} can be split into the product of two simpler
integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \bs{v} \, L(\bs{v}, d) \, p(\bs{v}) \, p(d\given\varpi)
\end{equation}
In either case, the marginalization over true 3-space velocity can be performed
analytically; we derive the relevant expressions in
\sectionname~\ref{sec:appendix}. After marginalizing over velocity, the
likelihood integrands only depend on distance; we numerically compute the
integrals over the true distances to each star in a pair, $d_1,d_2$, using Monte
Carlo integration with $K$ samples from the distance posterior pdfs:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

@@

\todo{clean this up}

>>>>

% to-do list
% ----------
% - check for TODO's and references to APW, HOGG, SMOH
% style notes
% -----------
% - This file generates by Makefile; don't be typing ``pdflatex'' or some bullshit.
% - Line break between sentences to make the git diffs readable.
% - Use \, as a multiply operator.
% - Reserve () for function arguments; use [] or {} for outer shit.
% - Use \sectionname not Section, \figname not Figure, \documentname not Article or Paper or paper.

@@

% packages
\definecolor{cbblue}{HTML}{3182bd}
\usepackage{microtype}  % ALWAYS!
\usepackage{amsmath}
\hypersetup{backref,breaklinks,colorlinks,urlcolor=cbblue,linkcolor=cbblue,citecolor=black}
\usepackage{dsfont}

@@

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\msun}{{\rm M}_\odot}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}
\newcommand{\eye}{\mathds{1}}

@@

The abundance of pairs of stars with small velocity difference in the
high-precsision sample of \figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at separations as large
as a few parsecs. Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances
    $\matrx{C}$.
  \item We assume that the 3-space velocities of a given pair of stars
    $(\bs{v}_i, \bs{v}_j)$ in the \tgas\ sample (relative to the solar system
    barycenter) are either (1) co-moving with velocity drawn from a velocity
    prior $p(\bs{v})$ and  velocity difference drawn from a zero-mean Gaussian
    with velocity dispersion $s$, or (2) individually drawn from a velocity
    prior $p(\bs{v})$.
\end{itemize}

Under these assumptions, the likelihood of a proper motion measurement for a
star given true distance, $d$, and true tangential velocity, $\bs{v}^\perp =
(\begin{array}[t]{c c} v_\alpha & v_\delta\end{array})^\mathsf{T}$ is
\begin{align}
  L(\bs{v}, d, s^2) &=
    \left[\det\left(\frac{\tilde{\matrx{C}}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{\mu} - \bs{x}_\theta \right)} \,
    \tilde{\matrx{C}}^{-1} \,
    \left(\bs{\mu} - \bs{x}_\theta \right) \right] \\
  \bs{x}_\theta &= d^{-1} \, \bs{v}^\perp
\end{align}
where the tangential velocity $\bs{v}^\perp$ is related to the 3-space velocity
$\bs{v}$ through projection onto a tangent plane at a given star's sky position
$(\alpha, \delta)$
\begin{align}
  \bs{v}^\perp &= \matrx{M}\,\bs{v} \\
  & = \left(
      \begin{array}{c c c}
        -\sin\alpha & \cos\alpha & 0 \\
        -\sin\delta \, \cos\alpha & -\sin\delta \, \sin\alpha & \cos\delta
      \end{array}
    \right) \,
    \left(\begin{array}{c} v_x \\ v_y \\ v_z \end{array}\right)
\end{align}
and the modified covariance matrix $\tilde{\matrx{C}}$ is
\begin{equation}
  \tilde{\matrx{C}} = \matrx{C} + \eye\,s^2
\end{equation}

We compute the fully marginalized likelihood (FML) that a given pair of stars
has the same 3-space velocity with a small velocity difference drawn from a
zero-mean Gaussian with variance $s^2$ (hypothesis 1, $\mathcal{L}_1$), and
the FML of the stars having different 3-space velocities (hypothesis 2,
$\mathcal{L}_2$).
We use the FML ratio $\mathcal{L}_1/\mathcal{L}_2$ as a scalar for selecting
candidate co-moving stars, as described below in more detail.
To compute these FMLs, the likelihood functions $L_i, L_j$ are marginalized
over true 3-space velocity and distance for each star in the pair $(i,j)$.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i, s^2) \,
    L_j(\bs{v}, d_j, s^2) \,
    p(\bs{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i, 0) \,
    L_j(\bs{v}_j, d_j, 0) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j). \label{eq:hyp2}
\end{align}
The marginalization integral for hypothesis 2 can be split into the product of
two simpler integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \bs{v} \, L(\bs{v}, d, 0) \, p(\bs{v}) \, p(d\given\varpi)
\end{equation}
If the velocity prior $p(\bs{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
\todo{SMOH: is this what we do?}
We use an equal-weight mixture of three isotropic, zero-mean Gaussian
distributions
\begin{equation}
  p(\bs{v}) = \frac{1}{3} \, \sum_{m=1}^3 \, \mathcal{N}(0, \sigma_m^2)
\end{equation}
with velocity dispersions $(\sigma_1, \sigma_2, \sigma_3) = (8, 32, 128)~\kms$
meant to represent young disk stars, old disk stars, and halo stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances to each
star in a pair, $d_1,d_2$, using Monte Carlo integration with $K$ samples from
the distance posterior pdfs:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
where $\tilde{L}(d)$ is the velocity-marginalized likelihood function.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

@@

\todo{APW will clean this up}
